---
title: "Linear models"
author: "Richard J. Telford"
date: "May 26, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Linear Modelling Exercises

1. Import the penguin data from the `palmerpenguin` package
```{r library}
library(palmerpenguins)
```


2. Do big birds have big beaks? Fit a linear model between bill length and body mass, and interpret the coefficients.

```{r linear-model-bill-beak}
fit.lm <- lm(bill_length_mm~body_mass_g, data = penguins)

summary(fit.lm)

beaks <- ggplot(fit.lm, aes(x=bill_length_mm, y=body_mass_g)) +
  geom_point()

print(beaks)

beta0 <- coefficients(fit.lm)[1]
beta1 <- coefficients(fit.lm)[2]
```



3. Examine the diagnostics plots. Is everything OK?

```{r diagnostic-plots}
par(mfrow=c(2,2))
plot(fit.lm)
```

Normal Q-Q
supposed to be a line; do residuals fit to a normal distribution? yes

Residuals vs fitted
Check for *outliers, variations in the mean res.*
Win: the line is totally flat, like it's supposed to be

Scale-Location
want a flat line - not too flat but i guess it's okey



3. Calculate the residual sum of squares (find sum the squared difference between the estimates calculated from the coefficients and the observed values).

```{r rss}
res_sum_sq <- sum((penguins$bill_length_mm - 
                    (beta0 + 
                       beta1 * penguins$body_mass_g))^2, na.rm=TRUE)

res_sum_sq
```


4. Recalculate the Residual sum of squares using different values for the coefficients (in the formula yi = Beta + Beta xi + εi).

```{r rss-new}
res_sum_sq2 <- sum((penguins$bill_length_mm - 
                    ((beta0+4) + 
                       (beta1-0.001) * penguins$body_mass_g))^2, na.rm=TRUE)

res_sum_sq2
```


5. Is the new RSS larger or smaller?

the new RSS is larger, since the linear model fitted to the data should show the least squared distance between the residuals and the fitted line

6. Estimate the slope using covariance and variance.

```{r}

#store the vectors as variables
x <- penguins$body_mass_g
y <- penguins$bill_length_mm

   
#slope using cov and var
Sxy <- cov(x=x, y=y, use="complete.obs")
Sx <- var(x=x, use="complete.obs")

slope_b1 <- Sxy / (sqrt((Sx^2)*Sx^2))
slope_b1
```


7. Do you get the same result? 

Unsure - the same result as what?

8. Add species to the model as a predictor.

```{r}
lm.fit2 <- lm(bill_length_mm ~ body_mass_g+species, data=penguins)
lm.fit2
```

9. Does species improve the model?

Yes, the species predictor does improve the model.
(Need to ask what the coefficients under the species' names mean. correlations?)

10. How should the coefficients be interpreted?

